Comprehensive Algorithmic Resilience for \\Numeric Applications

### 2013-10-16 paper.tex
First major pass through paper by Greg.

%\documentclass[prodmode,acmtecs]{acmconf} % Aptara syntax
\documentclass{sig-alternate}
\usepackage{url}
\usepackage{ulem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{color}
\usepackage[ruled]{algorithm2e}

@@

\title{Comprehensive Algorithmic Resilience for \\Numeric Applications}
%\author{Sui Chen,
%\affil{Louisiana State University}
%Greg Bronevetsky, Marc Casas,
%\affil{Lawrence Livermore National Laboratory}
%Lu Peng
%\affil{Louisiana State University}
%}

@@

\maketitle

\begin{abstract}
As High-Performance Computing (HPC) systems become larger and more complex and the feature sizes of their electronic components grow smaller, the systems become increasingly vulnerable to soft faults.
Since such faults may cause application crashes or may silently corrupt application results it is necessary to develop techniques to protect applications from such errors at a low cost in performance and power.
\greg{Since general-purpose resilience mechanisms, such as replication, can be very expensive for general applications it is imperative to develop ways to tailor resilience techniques to the needs and properties of each application.}

\greg{Extensive ongoing work on algorithmic techniques to detect and tolerate errors in common scientific sub-routines and kernels has shown that such techniques can be very effective at protecting the numeric properties of these code regions.
However, the practical utility of these techniques has been limited by the fact that real scientific applications (i) are significantly more complex than these simple kernels and (i) are vulnerable to a wider range of errors than most algorithmic checkers focus on.
To make scientific applications resilient to soft errors it is necessary to comprehensively combine resilience techniques that detect all the major types of errors that can affect an application and provide techniques that can tolerate them.
Further, the techniques applied to a given application must be tailored to an application's needs and configured to provide the performance for each input, hardware error rate and application user's accuracy tolerance.
This paper presents a case study of the design of comprehensive resilience mechanisms for four(\textbf{???}) different applications.
In addition to exploring the space of mechanism design and configuration we project our results to massively parallel applications and show how application performance and resilience configuration choices vary with error frequency, application scale and accuracy goals.
}

@@

The increasing size and complexity of HPC systems is making them increasingly vulnerable to soft faults.
Soft faults are transient corruptions of circuit state caused by physical phenomena such as strikes by neutrons or alpha particles~\cite{baumann:2005, asciQSER:2005} or thermal electrical noise~\cite{therm_noise:2007}.
They can affect the state of processor latches and registers and may cause the application to crash or worse, to silently return incorrect results~\cite{mpi_ser:reed:2004}.
As the feature sizes of electronic circuits shrink, technology scaling will make soft errors a larger problem~\cite{err_scaling:2012} due to the fact that each circuit element will hold less charge and can thus be disrupted more easily.
These phenomena make it imperative to develop techniques to make HPC systems resilient to soft faults.

Resilience is a problem that must be addressed at all levels.
While materials science and circuit design techniques can be used to improve resilience, they come at a cost in reduced power efficiency and performance that can become prohibitive if processors need to be sufficiently reliable to build a large HPC system.
Techniques such as error correcting codes have been very effective at making memories and caches resilient to soft faults~\cite{mem_errors:2010}.
However, they are more expensive for protecting core-internal state such as latches and are significantly less effective for checking the correctness of computations.
Traditional approaches like cross-core or node replication~\cite{rmpi:2011, dyn_cmp_repl:2007} are expensive in their use of computing resources and power and their long error detection latencies require processor state to be frequently checkpointed.
Processor designs that incorporate instruction replication~\cite{repl_smt:2000} offer fine-grained error detection and rollback but require more power as well as novel hardware features that are not likely to be included in the commodity processors used in HPC systems for cost reduction reasons.

\greg{The limitations of automated resilience techniques has motivated the need for algorithmic techniques that can detect and tolerate errors by taking advantage of application-specific properties and invariants~\cite{amg_abft:2012, robustification:2010, abft:1984}.
Although these techniques can be very efficient, most research work on such techniques has primarily focused on how their algorithmic invariants can be leveraged to enable accurate error detection.
While this work is crucial, it addresses only a portion of the full challenge of making scientific applications resilient to soft faults.
A comprehensive resilience strategy must address three areas.
First, it must deal all the possible application-level manifestations of soft faults including numeric errors (the focus of traditional algorithmic checkers), as well as corruptions in data structures and other application state and computations.
Second, it must address tolerance to soft faults in addition to their detection.
Finally, given the many ways in which resilience techniques can be configured (e.g. different detection tolerances for an algorithmic checker) it must provide ways to choose the best configuration for each system and for each user's performance and accuracy needs.}

\greg{
This paper presents a case study of how existing resilience techniques can be combined to make a usable resilience solution for scientific applications.
Within the context of three real applications we consider various combinations of algorithmic error detection, light-weight checkpoint restart to tolerate detected errors, as well as replication of pointers within data structures.
We focus on the Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}.
Each application requires different resilience techniques, each of which can be configured in different ways.
We use extensive fault injection experiments to measure the performance/resilience properties of a large space of resilience configuration within each application.
For each configuration we measure the slowdown induced by the execution of resilience logic and roll-backs due to error detections, as well as the distribution of magnitudes of error in the application's output.
We then demonstrate how these measurements can be used to configure the resilience mechanisms used with each application to minimize overhead while meeting the user's accuracy requirements.
Since in practice users do not have a clear idea of the level of accuracy needed from an algorithm's results, we demonstrate a novel alternative method to specify accuracy requirements as the level noise in the application's input required to deviate the output by a given magnitude (i.e. similar to backward error analysis).
Finally, we extrapolate from the small-scale application runs considered in our fault injection campaign (the many runs required to cover the large space of resilience configurations limited the length of each run) to application runs with 1 million CPU hours via three different models of error propagation.
Our results demonstrate that when applications use all the relevant resilience methods, configured for maximum efficiency, the overhead of resilient execution on expected Exascale designs is low.

@@

The paper is organized as follows.
Section~\ref{sec:soft_faults} discusses the problem of soft errors and presents the fault injection model we use to simulate them in our experiments.
The applications to which we apply our resilience methodology are described in Section~\ref{sec:apps}.
Section~\ref{sec:res_tech} summarizes the major routines of these applications and describes how these resilience techniques are applied to them.
Further, it experimentally evaluates the resilience techniques, describing their overheads and the accuracy levels of the protected routines at a wide range of possible error rates.
The effectiveness of our methodology for protecting the full applications is evaluated in Section~\ref{sec:eval}.

@@

Cosmic rays or alpha particles hitting the computer circuits and accumulating a sufficient amount of energy may invert the device logic states from "0" to "1" or from "1" to "0"~\cite{Ziegler:1996:TCR:226354.226356,baumann:2005}.
These inversions are termed as soft errors or soft faults.
The shrinking processor and memory feature size and the lower threshold voltage coming with relentless manufacturing technology scaling make modern high performance computers are highly vulnerable to soft errors which propagate as the program is being executed.
Prior studies have demonstrated that soft errors occurring in all the important bits that are required for Architecturally Correct Execution (ACE)~\cite{avf:2003} eventually change the program outcome, i.e., manifest errors to the application.
The ACE bits include critical information to keep the program logic such as the program counter register for a committed instruction.
In the past decades, soft errors have brought significant losses on commercial servers and high performance computers such as the recall of Sun servers due to memory errors~\cite{baumann:2005} or the cache corruptions in the ASCI Q system~\cite{asciQSER:2005}.

@@

We use a fault injection infrastructure~\cite{relax:2010} that simulates the architecture-level effects of soft faults.
It works by compiling application into the Low Level Virtual Machine (LLVM) typed bytecode~\cite{LLVM}.
This bytecode is transformed to allow a bit flip in each instruction's output with some probability $P$.
The time between injections is thus exponentially distributed with period $\frac{1}{P}$.
Our fault injector is a high-level approximation of the physical corruptions caused by soft errors that strikes a balance between model accuracy and cost, as do related approaches~\cite{fault_injection:iyer:1997, avf:2003}.
Alternative models, such as injection into gate-level processor models~\cite{sesee:2004} or neutron beam irradiation~\cite{freq_dep:irom:2004} are possible but their significantly higher cost makes them impractical for conducting large-scale fault injection campaigns such as the ones reported in this paper.

\sui{As a metric for quantifying errors in output caused by soft faults, we perturb inputs to numeric applications ...}

@@

We evaluate our resilience methodology by applying it to the following three applications.
Since they spend most of their execution time in library routines, we focus our resilience efforts on protecting these routines, which are described in Section~\ref{sec:res_tech:err_det:algo}.

%\greg{This section needs to include an experimental breakdown of where the routines spent their time. We also need to describe the types of inputs we'll be giving them in our experiments.}

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
The ADDR method solves under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing the function $f(x)$.
This is an iterative algorithm uses 64-bit precision and that spends most of its time in the following linear algebra operations, matrix-matrix multiplication, matrix-vector multiplication, rank-k update and Cholesky factorization, using the implementations of these routines from the GNU Scientific Library~\cite{gsl:2011}.
We focus on a specific application that uses the ADDR method to solve Lasso problems, which are defined as minimizing the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$.
It is parallelized using MPI and our experiments focus on 4-processors runs of ADDR which computes linear systems of size $\{800, 1200, ..., 3600\} \times 500$ (800 to 3600 measurements with 500 variables each) and in the experiments below are denoted $S800$, $S1200$, ..., $S3600$.
The values in $A$ and $b$ are generated by sampling a normal distribution with a mean of 0.
According to our experiments, the running time of this solver is not dependent on the kind of distribution the inputs are sampled from.

@@

DRC is a sequential program generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners.
The inputs are stored in Pulse Code Modulation (PCM) format, which is an array of 32-bit floating point numbers representing the samples at each sampling time step and the computations are performed using 32-bit precision.
Although the algorithm is divided into many phases the most computationally intensive step consists of the GSL's implementation of the Fast Fourier Transform and a DRC-internal implementation of Finite Impulse Response filter generation.
This program accepts user-controllable configurations to decide the width and number of iterations of FFT transforms it uses to perform various passes on the input file, such as windowing, deconvolution, pre- and post-filtering.
The input used in this paper is an audio file of size 768 kilobytes, which is sampled at the rate of 30Khz, 40Khz, ..., 90Khz.

\subsection{Hattrick}
\label{sec:apps:hattrick}
Hattrick is a sequential application simulates the motion of bodies under the effects of gravity using 64-bit precision.
It is designed to help discover extra-solar planets by inferring their existence from Transit Timing Variations, where the difference between the times when a planet passes in front of its host star is used to infer the existence and properties of other planets in the system.
This program spends most of its execution time in the GSL solver for Ordinary Differential Equations to solve the system's equations of motion.
A given input is described using three parameters: $P$=number of planets, $T$=amount of time to simulate, and $A$=the accuracy target.
In our experiments we considered the following four inputs: $P2T2090A15$, $P2T3090A15$, $P2T4090A15$ and $P3T2090A11$, where an $A15$ and $A11$ denote accuracy targets of $1e-15$ and $1e-11$, respectively.

@@

In this paper we consider a comprehensive resilience strategy that combines various types of error detection and recovery.
Error detection techniques are discussed in Section~\ref{sec:res_tech:err_det}, with Section~\ref{sec:res_tech:err_det:algo} focusing on algorithmic techniques to detect errors in the numeric portions of application state, while Section~\ref{sec:res_tech:err_det:mem} focuses on protection of data structures.
Section~\ref{sec:res_tech:cr} then considers recovery techniques based on both checkpoint-restart, which offers lower-overhead but higher-latency recovery, as well as replication of key pointers, which reduces performance but enables fast recovery.
Finally, Section~\ref{sec:res_tech:eval} presents an experimental evaluation of the performance and accuracy properties of the resilient versions of the major routines used by our target applications, under a wide range of inputs and error injection rates.

@@

%\sui{Performance experiments inserted here: running algorithmic checkers across different input sizes and fault rates: 62x62 to 500x500 matrices for MM, RK, CD, MV checkers; FFT sizes of $2^8$ to $2^{24}$ for the FFT checker; Error bars should apply here, especially for the Cholesky Decomposition checker}

Each algorithm used by the above application was enhanced with an algorithm-specific checker that validated its results by exploiting some algorithmic identity.
Each checker computes some aggregate property of the algorithm's results using two different algorithms to produce vectors $v$ and $v'$.
If their relative difference by more than some threshold $\tau$, the checker declares that an error has occurred.
The relative difference is computed as $\frac{\left|| v-v' \right||}{max(\left||v\right||, \left||v'\right||)}$.
Each error detection causes the routine's execution to be restarted using the rollback mechanism described in Section~\ref{sec:res_tech:cr}.

\vspace{-5pt}
\paragraph{Matrix-matrix multiplication (MMM)}
The operation $A \cdot B$ is checked using a matrix vector multiplication (MVM), via the identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector.
Since we set the check vector $x$ to be all 1s this is equivalent to both sides computing the sums of the columns of matrix $A \cdot B$ using a single MVM on the left and two MVMs on the right.
Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
This check is efficient because the checker is asymptotically faster than MMM, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.

\vspace{-5pt}
\paragraph{Matrix-vector multiplication (MVM)}
The MVM algorithm $Ax=b$ is checked using a similar identity $(x^TA)x = x^T(Ax) = x^Tb$.
We again set $x$ to be the vector of 1's, making the product $x^TA$ equivalent to sum of the columns of matrix $A$.
Although the computation of $x^TA$ involves additions rather than multiplications, the complexity of computing $x^TA$ is $O(n^2)$, the same as the original MVM operation.
%While this cost can be reduced significantly by computing $x^TA$ once and reusing it in subsequent multiplications by the same matrix, we did not consider it in this paper because MVM took up little application execution time.
However, for applications such as ADDR where the MVM is applied to same matrix many times with different vectors, the column sums can be reused, amortizing the cost of this computation.
Our experimental evaluation below thus focuses on the overhead of the $x^T(Ax)$ computation, which has $O(n)$ complexity.

\vspace{-5pt}
\paragraph{Symmetric Rank-K update (SYRK)}
The SYRK update is a special case of matrix-matrix multiplication $B = \alpha A \cdot A^T + \beta B$, where an $n \times k$ matrix $A$ is multiplied by itself to produce an $n \times n$ matrix with rank $k$, which is incremented into the $n \times n$ matrix $B$.
Its results are checked using the same algorithm as MMM by checking the identity $B x = \alpha A \cdot (A^Tx) + \beta B x$, where $x$ is the vector of all 1's.
This check runs in time $O(n^2)$, as compared to the $O(n^3)$ cost of SYRK.

\vspace{-5pt}
\paragraph{Cholesky Decomposition (CD)}
In this decomposition matrix $A$ is decomposed into $L \cdot L^T$ where $L$ is a lower-triangular matrix with a positive diagonal.
This routine is checked just like MMM by checking the identity $Ax = L \cdot (L^T x)$, which is cheaper than the $O(n^3)$ complexity of the deterministic CD algorithm.
The GSL implements an iterative algorithm that runs faster than $O(n^3)$ time but as shown in the experiments below, it is significantly slower than our checker.

\vspace{-5pt}
\paragraph{Fast Fourier Transform (FFT)}
The FFT algorithm decomposes a function into a sum of sine waves of different frequencies: $f(x) = \sum_{n=0}^{N-1} x_n e^{-i2\pi k n / N}$ for some constant $k$.
We check the results of FFT by using Parseval's theorem, which states that $\sum_{n=0}^{N-1} \left| x[n] \right|^2 = \frac{1}{N} \sum_{k=0}^{N-1} \left| X[k] \right|^2$, where $x$ is the original function and $X$ is its transform.
Intuitively it means that the energy of the original function is preserved by the transform.
GSL implements optimized radix 2, 3, 4, 5, 6 and 7 subtransform routines as well as a general radix-N FFT routine. The subtransforms are optimized and consume $O(n log(n))$ time and the general one consumes $O(n^2)$ time. Therefore, GSL has a scheduling algorithm that factorizes the length of the input FFT, running the subtransforms whenever possible, using the fallback general radix-N routine only for the remaining factors.
The theorem applies to FFT of any radix and provides an $O(n)$ time check for the $O(n log(n))$ or $O(n^2)$ FFT algorithms.

>>>>

%\documentclass[prodmode,acmtecs]{acmconf} % Aptara syntax
\documentclass[10pt, conference, compsocconf]{IEEEtran}
\usepackage{url}
\usepackage{ulem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{color}
\usepackage[ruled]{algorithm2e}
\usepackage{multirow}

@@

\title{Comprehensive Algorithmic Resilience for \\Numeric Applications}
%\author{Sui Chen,
%\affil{Louisiana State University}
%Greg Bronevetsky,
%\affil{Lawrence Livermore National Laboratory}
%Marc Casas,
%\affil{Barcelona Supercomputing Center}
%Lu Peng
%\affil{Louisiana State University}}

\author{\IEEEauthorblockN{Sui Chen}
\IEEEauthorblockA{Louisiana State \\ University}
\and
\IEEEauthorblockN{Greg Bronevetsky}
\IEEEauthorblockA{Lawrence Livermore \\ National Laboratory}
\and
\IEEEauthorblockN{Marc Casas-Guix}
\IEEEauthorblockA{Barcelona Supercomputing \\ Center}
\and
\IEEEauthorblockN{Lu Peng}
\IEEEauthorblockA{Louisiana State \\ University}
}

@@

\maketitle
\begin{abstract}

As High-Performance Computing (HPC) systems become larger and more complex and their circuit feature sizes shrink, the systems become increasingly vulnerable to soft faults.
Since such faults may cause application crashes or may silently corrupt application results it is necessary to develop techniques to protect applications from such errors at a low cost in performance and power.
Since general-purpose resilience mechanisms, such as replication, can be very expensive it is imperative to develop ways to tailor resilience techniques to each application's properties and the needs of its users.

Extensive ongoing work on algorithmic techniques to detect and tolerate errors in common scientific sub-routines and kernels has proven their effectiveness.
However, their practical utility has been limited by the fact that real scientific applications (i) are significantly more complex than these simple kernels and (i) are vulnerable to a wider range of errors than most algorithmic checkers focus on.
To make scientific applications resilient to soft errors it is necessary to comprehensively combine resilience techniques that tolerate \emph{all} the major types of application errors.
Further, each technique must be tailored to an application's needs and configured to provide the performance for each input, hardware error rate and application user's accuracy tolerance.
This paper presents a case study of the design of comprehensive resilience mechanisms for three application domains.
We describe the combinations of techniques required to make a single application from each domain resilient and quantify their performance and resilience on small-scale inputs.
Further, we use these experiments to predict the behavior of large-scale instances (1 million CPU hours) of applications in each domain when using the most efficient application configuration for each fault rate and user-specified accuracy requirement.

@@

The increasing size and complexity of HPC systems is making them increasingly vulnerable to soft faults.
Soft faults are transient corruptions of circuit state caused by physical phenomena such as strikes by neutrons or alpha particles~\cite{baumann:2005, asciQSER:2005} or thermal electrical noise~\cite{therm_noise:2007}.
They can affect processor latches and registers and may cause the application to crash or worse, to silently return incorrect results~\cite{mpi_ser:reed:2004}.
As circuit feature sizes shrink, technology scaling will make soft faults a larger problem~\cite{err_scaling:2012}, %due to the fact that each circuit element will hold less charge and can thus be disrupted more easily.
which makes it imperative to develop techniques to make HPC systems resilient to soft faults.

Resilience must be addressed at all system levels.
Materials science and circuit design techniques can improve resilience but at the cost of reduced power efficiency and performance that can become prohibitive at processor reliability levels needed to build a large HPC system.
Error correcting codes are very effective at making memories and caches resilient to soft faults~\cite{mem_errors:2010} but are much more expensive for latches and significantly less effective for computational circuits.
Replication across cores or nodes~\cite{rmpi:2011, dyn_cmp_repl:2007} more than double resource requirements and their long error detection latencies require frequent checkpointing.
Processor designs that support instruction replication~\cite{repl_smt:2000} offer fine-grained error detection and rollback but require more power as well as novel hardware features unlikely to be included in commodity processors.

The limitations of automated resilience techniques motivates work on algorithmic detection and tolerance techniques by leveraging application-specific properties and invariants~\cite{amg_abft:2012, robustification:2010, abft:1984}.
Although very efficient, these techniques are not general and most research has focused on algorithmic resilience for numeric kernels.
While this is crucial, it addresses only a portion of the full challenge of making scientific applications resilient to soft faults.
A comprehensive resilience strategy must address three areas.
First, it must deal all the possible application-level manifestations of soft faults including numeric errors (focus of traditional algorithmic checkers) as well as corruptions in data structures and other state and computations.
Second, it must address tolerance to soft faults in addition to their detection.
Finally, given the many ways to configure resilience techniques (e.g. detection tolerances) it must provide ways to choose the best configuration for each system and for each user's performance and accuracy needs.

This paper presents a case study of how existing resilience techniques can be combined to make an efficient resilience solution for large-scale scientific applications.
Our analysis focuses on linear solver, signal processing and n-body simulation applications, using the following applications as representatives: Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}, respectively.
We use small-scale runs of each application on multiple inputs to comprehensively measure their performance and resilience properties when protected by various various combinations of algorithmic error detection, light-weight checkpoint restart (tolerates detected errors), as well as pointer replication.
We then extrapolate from these small-scale experiments to predict the performance and resilience properties of large-scale runs (1 million CPU hours) of applications within each domain.
This is done by modeling large-scale runs as dependence graphs that connect small-scale runs into parallel reductions, sequential chains or a combination of both.
Our analysis explicitly accounts for the accuracy that users may require from applications in terms of a bound on the Root Mean Squared (RMS) Error of the results as well as a bound on the probability that this target will not be met.
To simplify the process of choosing the appropriate error bound we also provide a characterization of application output error in terms of the amount of measurement noise in application inputs that may produce the same deviation in application outputs (similar to backward error analysis).
This makes it possible for application users to rationally trade off the accuracy of application results against the cost to achieve it (e.g. power, time or system purchase price).

%Each application requires different resilience techniques, each of which can be configured in different ways.
%We use extensive fault injection experiments to measure the performance/resilience properties of a large space of resilience configuration within each application.
%For each application configuration we measure the slowdown induced by the execution of resilience logic and roll-backs due to error detections, as well as the distribution of magnitudes of error in the application's output.
%We then demonstrate how these measurements can be used to configure the resilience mechanisms used with each application to minimize overhead while meeting the user's accuracy requirements.
%Since in practice users do not have a clear idea of the level of accuracy needed from an algorithm's results, we demonstrate a novel alternative method to specify accuracy requirements as the level noise in the application's input required to deviate the output by a given magnitude (i.e. similar to backward error analysis).
%Finally, we extrapolate from the small-scale application runs considered in our fault injection campaign (the many runs required to cover the large space of resilience configurations limited the length of each run) to application runs with 1 million CPU hours via three different models of error propagation.
%Our results demonstrate that when applications use all the relevant resilience methods, configured for maximum efficiency, the overhead of resilient execution on expected Exascale designs is low.

@@

The paper is organized as follows.
Section~\ref{sec:soft_faults} discusses the problem of soft errors and presents the fault injection model we use to simulate them in our experiments.
The applications we focus on are described in Section~\ref{sec:apps}.
Section~\ref{sec:res_tech} summarizes the major routines of these applications and describes the resilience techniques that are applied to them.
It further details our methodology to evaluate application performance and resilience in the face of faults.
Finally, Section~\ref{sec:eval} describes our procedure to use small-scale runs of our three target applications to predict the performance and resilience properties of large-scale runs of applications within their respective domains.

@@

The states of logic devices can be corrupted due to fluctuations in circuit voltages or strikes by charged particles from chip packaging or due to cosmic rays~\cite{Ziegler:1996:TCR:226354.226356,baumann:2005}.
These ``soft faults'' are expected to become increasingly common as circuit scaling further shrinks circuit feature size and the threshold voltage.
Prior studies have demonstrated that soft errors in processor state required for Architecturally Correct Execution (ACE)~\cite{avf:2003,mpi_ser:reed:2004} eventually change the program outcome.
%The ACE bits include critical information to keep the program logic such as the program counter register for a committed instruction.
In the past decades, soft errors have brought significant losses on commercial servers and high performance computers such as the recall of Sun servers due to memory errors~\cite{baumann:2005} or the cache corruptions in the ASCI Q system~\cite{asciQSER:2005}.

@@

We simulate soft faults using a fault injection infrastructure~\cite{relax:2010} that simulates the architecture-level effects of soft faults.
It works by compiling application into the Low Level Virtual Machine (LLVM) typed bytecode~\cite{LLVM} and transforming it to allow bits in instruction outputs to be flipped with some probability.
%The time between injections is thus exponentially distributed with period $\frac{1}{P}$.
Our fault injector is a high-level approximation of the physical corruptions caused by soft errors that strikes a balance between model accuracy and cost, as do related approaches~\cite{fault_injection:iyer:1997, avf:2003}.
Alternative models, such as injection into gate-level processor models~\cite{sesee:2004} or neutron beam irradiation~\cite{freq_dep:irom:2004} are possible but too expensive for large-scale fault injection campaigns such as the ones in this paper.

@@

We evaluate our resilience methodology by applying it to the following three applications.
Since they spend most of their execution time in library routines, we focus our resilience efforts on protecting these routines, as described in Section~\ref{sec:res_tech:err_det:algo}.

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
ADDR is an implementation of an iterative method for solving under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing some function $f(x)$.
It represents the linear solver application domain.
It uses 64-bit precision and spends most of its time in the following linear algebra operations from the GNU Scientific Library~\cite{gsl:2011}: matrix-matrix multiplication (MMM), matrix-vector multiplication (MVM), rank-k update (RK) and Cholesky decomposition (CD).
We use ADDR to solve Lasso problems, where the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$ is minimized.
It is parallelized using MPI and our experiments focus on 4-processors runs with linear systems of size $\{800, 1200, ..., 3600\} \times 500$ as input (denoted $S800$, ..., $S3600$ below).%(800 to 3600 measurements with 500 variables each) and in the experiments below are denoted $S800$, $S1200$, ..., $S3600$.
The values in $A$ and $b$ are generated by sampling a normal distribution with a mean of 0. \greg{What's the $\sigma$?}
According to our experiments, the running time of this solver is not dependent on the distribution the inputs are sampled from.

@@

DRC is a sequential program that generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners.
It represents the signal processing application domain.
DRC's inputs are stored in Pulse Code Modulation (PCM) format, which is an array of 32-bit floating point numbers representing the samples at each time step and the computations are performed using 32-bit precision.
Most execution time is spent in the GSL implementation of Fast Fourier Transform (FFT) and a DRC-internal implementation of Finite Impulse Response (FIR) filter generation.
%This program accepts user-controllable configurations to decide the width and number of iterations of FFT transforms it uses to perform various passes on the input file, such as windowing, deconvolution, pre- and post-filtering.
The input used in this paper is an audio file of size 768 kilobytes, which is sampled at the rate of 30Khz, 40Khz, ..., 90Khz.

\subsection{Hattrick}
\label{sec:apps:hattrick}
Hattrick is a sequential application that simulates the motion of bodies under the effects of gravity to help discover extra-solar planets by inferring their existence from Transit Timing Variations.
%, where the difference between the times when a planet passes in front of its host star is used to infer the existence and properties of other planets in the system.
It represents the n-body simulation application domain.
Hattrick uses 64-bit precision and spends most of its execution time in the GSL solver for Ordinary Differential Equations to solve the system's equations of motion using the Runge-Kutta method (RK).
A given input is described using three parameters: $P$=number of planets, $T$=amount of time to simulate, and $A$=the accuracy target.
In our experiments we considered the following four inputs: $P2T2090A15$, $P2T3090A15$, $P2T4090A15$ and $P3T2090A11$, where an $A15$ and $A11$ denote accuracy targets of $1e-15$ and $1e-11$, respectively.
\greg{Does it take transit timings as inputs? What are you perturbing?}

@@

Our case-study focuses on a comprehensive resilience strategy that combines various types of error detection and recovery.
Algorithmic error detection techniques of numeric state are discussed in Section~\ref{sec:res_tech:err_det}, while Section~\ref{sec:res_tech:pointers} focuses on protection of data structure pointers via replication.
Section~\ref{sec:res_tech:cr} then considers recovery techniques based on both checkpoint-restart.%, which offers lower-overhead but higher-latency recovery, as well as replication of key pointers, which reduces performance but enables fast recovery.
Finally, Section~\ref{sec:res_tech:eval} describes our methodology to analyze the resilience and performance properties of these mechanisms in the above applications across a broad range of configurations.
This information is used in Section~\ref{sec:eval} to configure the resilience methods for modeled large-scale application runs to maximize the high performance across a broad range of fault rates and while satisfying user accuracy requirements.

@@

Algorithms used by the above application were enhanced with checkers of their algorithmic invariants.
Each checker computes some aggregate property of the algorithm's output in two different ways to produce values $v$ and $v'$.
An error is detected if their relative difference $\frac{\left|| v-v' \right||}{max(\left||v\right||, \left||v'\right||)}$ is larger than some threshold $\tau$.
Each error detection causes the routine's execution to be restarted using the rollback mechanism described in Section~\ref{sec:res_tech:cr}.

%\vspace{-5pt}
\paragraph{Matrix-matrix multiplication (MMM)}
$A \cdot B$ is checked using a matrix vector multiplication, via the identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector (we use the vector of all 1s).
%Since we set the check vector $x$ to be all 1s this is equivalent to both sides computing the sums of the columns of matrix $A \cdot B$ using a single MVM on the left and two MVMs on the right.
%Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
The checker is asymptotically faster than MMM, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.

%\vspace{-5pt}
\paragraph{Symmetric Rank-K update (SYRK)}
Since this is a special case of MMM where $B = \alpha A \cdot A^T + \beta B$, %where an $n \times k$ matrix $A$ is multiplied by itself to produce an $n \times n$ matrix with rank $k$, which is incremented into the $n \times n$ matrix $B$.
%Its results are checked using the same algorithm as MMM by checking the identity $B x = \alpha A \cdot (A^Tx) + \beta B x$, where $x$ is the vector of all 1's.
%This check runs in time $O(n^2)$, as compared to the $O(n^3)$ cost of SYRK.
we use the same check.

%\vspace{-5pt}
\paragraph{Matrix-vector multiplication (MVM)}
$Ax=b$ is checked using a similar identity $(x^TA)x = x^T(Ax) = x^Tb$.
%We again set $x$ to be the vector of 1's, making the product $x^TA$ equivalent to sum of the columns of matrix $A$.
The complexity of computing $x^TA$ is $O(n^2)$, the same as the original MVM but using additions rather than multiplications.
%While this cost can be reduced significantly by computing $x^TA$ once and reusing it in subsequent multiplications by the same matrix, we did not consider it in this paper because MVM took up little application execution time.
Since in ADDR MVM is applied many times to the same matrix with different vector, the vector $x^TA$ can be reused, amortizing the cost of computing it.
Our experiments below thus focus on the cost of computing $x^T(Ax)$, which has $O(n)$ complexity.

%\vspace{-5pt}
\paragraph{Cholesky Decomposition (CD)}
CD decomposes matrix $A$ into $L \cdot L^T$ where $L$ is lower-triangular with a positive diagonal.
It is checked via the identity $Ax = L \cdot (L^T x)$, which at $O(n^2)$ complexity is cheaper than the $O(n^3)$ complexity of the deterministic CD algorithm.
GSL implements an iterative algorithm that runs faster than $O(n^3)$ time but as our experiments show, our checker is still significantly faster.

%\vspace{-5pt}
\paragraph{Fast Fourier Transform (FFT)}
Decomposes a function into a sum of sine waves of different frequencies: $f(x) = \sum_{n=0}^{N-1} x_n e^{-i2\pi k n / N}$ for some constant $k$.
It is checked using Parseval's theorem: $\sum_{n=0}^{N-1} \left| x[n] \right|^2 = \frac{1}{N} \sum_{k=0}^{N-1} \left| X[k] \right|^2$, where $x$ is the original function and $X$ is its transform.
Intuitively it means that the energy of the original function is preserved by the transform.
%GSL implements optimized radix 2, 3, 4, 5, 6 and 7 subtransform routines as well as a general radix-N FFT routine. The subtransforms are optimized and consume $O(n log(n))$ time and the general one consumes $O(n^2)$ time. Therefore, GSL has a scheduling algorithm that factorizes the length of the input FFT, running the subtransforms whenever possible, using the fallback general radix-N routine only for the remaining factors.
%The theorem applies to FFT of any radix and provides an $O(n)$ time check for the $O(n log(n))$ or $O(n^2)$ FFT algorithms.
This check runs in $O(n)$ time, as compared to $O(n log(n))$ or $O(n^2)$ for the FFT algorithm itself (depends on the type of FFT problem).

%\vspace{-5pt}
\paragraph{Finite Impulse Response Filter Generation (FIR)}
The DRC FIR filter algorithm generates a series of samples over the function $sinc(x)=\frac{sin(x)}{x}$
%(\text{if}~x \ne 0)$ and $sinc(x)=1 (\text{if}~x=0)$
and modulates it with a Blackman window.
It is checked using the invariant $\int_{-\infty}^{\infty} sinc(x)dx = 1$, which is preserved by the Blackman window in most cases.
Computing the sum requires $O(n)$ additions, as compared to the $O(n)$ trigonometric function evaluations of the non-iterative FIR generation algorithm.
%As shown below, the overhead of this checker is very small (less than 5\%).
%Since this algorithm has only output and no input and therefore it does not need to have to be reinforced with input backup as other routines do.

%\vspace{-5pt}
\paragraph{Runge-Kutta PDE Solver (RK)}
A 4-th order Runge-Kutta method for solving Ordinary Differential Equations of the form $\frac{dy}{dx} = f(y, x)$.
It advances the variable $x$ by steps of size $h$ and computes the value $y$ at the next point $x+h$ using the derivative $\frac{dy}{dx}$ at $x$.
The GSL implementation of RK uses adaptive step-size control where it simultaneously uses two step sizes $h$ and $\frac{h}{2}$ (more precise).
If the difference between the two computations exceeds a threshold $\tau$, it switches to the smaller step size to maintain accuracy.% and resumes.
If it is smaller than $\frac{\tau}{2}$, the algorithm switches to a larger step size.
Because soft faults usually cause significant differences between the two computations, the adaptive step size control algorithm naturally tolerates them by temporarily increasing its step size and then reverting to the original step size when it observes that this is safe.
We thus protect the results of user-provided derivative functions within RK's steps using this algorithm.
%However, computations in user-provided function derivative functions are not checked since their invariants are not known.
